Coldplay vs Radiohead Danceability over Time

### 2024-10-10 paper/paper.qmd
Add files via upload

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Yunkyung Ko
  - Chenika Bukes
thanks: "Code and data are available at: [https://github.com/koyunkyung/spotify-API](https://github.com/koyunkyung/spotify-API)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(spotifyr)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

# References

>>>>

---
title: "Coldplay vs Radiohead Danceability over Time"
subtitle: "Coldplay reamins more danceable over the years"
author: 
  - Chenika Bukes
  - Khushaal Nandwani
  - Yongkyung Ko
thanks: "Code and data are available at: https://github.com/chenikabukes/RadioheadColdplay"
date: today
date-format: long
abstract: "This study explores the danceability of Coldplay and Radiohead over time by using data extracted from Spotify’s API. The results show that Coldplay has consistently maintained a higher level of danceability compared to Radiohead. These findings are relevant in understanding the musical evolution of both bands and their appeal to different types of listeners."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)

radiohead <- readRDS("../data/01-raw_data/radiohead.rds")
coldplay <- readRDS("../data/01-raw_data/coldplay.rds")
```

# Introduction
The estimand in this paper is the average danceability of tracks by both Coldplay and Radiohead over time. Danceability describes how suitable a track is for dancing based on a combination of tempo, rhythm stability, beat strength, and overall regularity. This paper shows how this metric changes for each band and discusses the implications.

Understanding how danceability evolves over time can provide insight into the bands' stylistic changes and the broader trends in the music industry. All analysis in this paper was completed using R [@citeR]. The remainder of this paper is structured as follows. Section @sec-data describes the data collection and preparation. Section @sec-results presents the findings using the tidyverse package [@tidyverse], and Section @sec-discussion discusses the implications of these results.

@@

The data used in this study was retrieved using the Spotify Web API and processed using the R package spotifyr [@Spotify]. The package allows bulk retrieval of an artist’s entire discography along with associated audio features, including danceability, tempo, energy, and others. We accessed both Coldplay's and Radiohead’s discographies and calculated the average danceability of their songs by year of release. The data consists of metadata about each track, including the release year, danceability score, and album information.

# Results {#sec-results}

In @fig-coldplay, we observe Coldplay's tracks exhibit relatively stable danceability over time, with scores ranging between 0.4 and 0.6, with slight increases toward the latest releases. Coldplay’s more consistent and upward trend aligns with their pop-rock style, which tends to favor rhythmically engaging and accessible music, contributing to a higher and more consistent danceability score. Radiohead’s danceability exhibits much more fluctuation, ranging from around 0.2 to 0.6. There is an increase during the early 2000s, but it dips again afterward. The lower and more fluctuating danceability for Radiohead reflects their tendency to diverge from mainstream pop structures.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| label: fig-coldplay
#| tbl-cap: "Danceability of Coldplay's and Radiohead's tracks over time"
#| message: false
#| fig-cap: "Danceability of Coldplay's and Radiohead's tracks over time"

rbind(coldplay, radiohead) |>
  select(artist_name, album_release_date, danceability) |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes( x = album_release_date, y = danceability, color = artist_name)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  facet_wrap(facets = vars(artist_name), dir = "v") +
  labs(
    x = "Album release date",
    y = "Danceability",
    color = "Artist"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

# Discussion {#sec-discussion}
Both bands show some increase in danceability in their later albums, though Coldplay's increase is more pronounced, reflecting their shift towards more pop-centric music.

@@

# Appendix
Out of interest we decided to explore the correlation between danceability and tempo. The below results do not show a significant relationship and thus were not included in the main findings. 
```{R}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-corrplot_r
#| fig-cap: "Danceability vs Tempo for Radiohead Tracks"
#| fig-width: 2.5
#| fig-height: 2

library(ggplot2)
library(dplyr)
radiohead <- readRDS("../data/01-raw_data/radiohead.rds")

#### Radiohead
ggplot(radiohead, aes(x = tempo, y = danceability)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "Tempo (BPM)",
       y = "Danceability") +
theme_minimal()

#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-corrplot_c
#| fig-cap: "Danceability vs Tempo for Coldplay Tracks"
#| fig-width: 2.5
#| fig-height: 2

library(ggplot2)
library(dplyr)
coldplay <- readRDS("../data/01-raw_data/coldplay.rds")

#### Coldplay
ggplot(coldplay, aes(x = tempo, y = danceability)) +
  geom_point(color = "green", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "Tempo (BPM)",
       y = "Danceability") +
  theme_minimal()
```
\newpage 
\newpage
# References
